A seamless, client-centric programming model for type safe web applications

### 2014-05-08 paper.tex
Stuff.

\begin{abstract}
We propose a new programming model for web applications which is (1)
seamless; one program and one language is used to produce code for
both client and server, (2) client-centric; the programmer takes the
viewpoint of the client that runs code on the server rather than the
other way around, (3) functional and type-safe; and (4) portable;
everything is implemented as a Haskell library that implicitly takes
care of all networking code. Our aim is to improve the painful and
error-prone experience of today's standard development methods,
in which clients and servers are coded in different languages and
communicate with each other using ad-hoc protocols.
We present the design of our library called Haste.App,
an example web application that uses it, and discuss
the implementation and the compiler technology on which it depends.
\end{abstract}

@@

For a more concrete example, let's say that we want to implement a simple
``chatbox'' component for a website, to allow visitors to discuss the site's
content in real time. Using mainstream development practices and recent
technologies such as WebSockets\ \cite{websockets}, we may come up with
something like the program in figure \ref{lst:javascript-client}.

@@

To run this example, an address and a port must be provided so that the client
knows which server to contact. There are several ways of doing this:
using the GHC plugin system, through Template Haskell or by slightly altering
how program entry points are treated in a compiler or wrapper script, to name
a few. A non-intrusive method when using the GHC/Haste compiler pair would
be to pass add \lstinline!-main-is setup! to both compilers' command line
and add the \lstinline!setup! function to the source code.
This will instruct the server binary to listen on the port 1111 when
started, and the client to attempt contact with that port on the local machine.

@@

The exact mechanism chosen to provide the host and port are implementation
specific, and will in the interest of clarity not be discussed further.

@@

\begin{code}
srvHello :: Server RcptList -> Server ()
srvHello remoteRcpts = do
  recipients <- remoteRcpts
  sid <- getSessionID
  liftIO $ do
    rcptMVar <- CC.newEmptyMVar
    CC.modifyMVar recipients $ \cs ->
      return ((sid, rcptMVar):cs ,())
\end{code}

@@

runClient    :: Client () -> App Done
liftServerIO :: IO a -> App (RemoteValue a)
remote       :: Remotable a
             => a -> App (Remote a)

@@

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!remote! & \parbox[t]{5cm}{Make a server side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!remote! function to
a serializable argument.} \\
\lstinline!use! & \parbox[t]{5cm}{Extract the contents of a
\lstinline!RemoteValue! produced by \lstinline!liftServerIO! on the server.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

The workings of the \lstinline!App! monad basically hinges on two abstract data
types: \lstinline!RemoteValue! and \lstinline!Remote!. \lstinline!RemoteValue!
is used to represent values which are available to the server but not to the
client. \lstinline!Remote! denotes functions which live on the server but can
be invoked remotely by the client. The implementation of these types and the
functions that operate on them differ between the client and the server.

\paragraph{Client side implementations}
We begin by looking at the client side implementation.

\begin{code}
data RemoteValue a = There
data Remote a = Remote CallID [JSON]
\end{code}

As values of the \lstinline!RemoteValue! type are not accessible to the client,
they are represented by a simple dummy value. The \lstinline!Remote! type is
slightly more useful, containing the ID of a remote function and a list of the
serialized arguments to be passed when invoking it. In essence, the
\lstinline!Remote! type is an explicit representation of a remote closure.
Such closures can be applied to values using the \lstinline!<.>! operator.

@@

The \lstinline!remote! function is used to bring server side functions into
scope on the client as \lstinline!Remote! functions. It is implemented using a
simple counter which keeps track of how many functions have been imported so
far and thus which ID to assign to the next remote function.

@@

\begin{code}
liftServerIO :: IO a -> App (RemoteValue a)
liftServerIO _ = App $ return There
\end{code}

As we can see, the implementation is as simple as can be. Since
\lstinline!RemoteValue!s are represented by a dummy value on the client, we
just return said value.

\paragraph{Server side implementations}
The server side representation of the \lstinline!RemoteValue! and
\lstinline!Remote! types are in a sense the opposites of their client side
counterparts.

\begin{code}
data Remote a = Remote
\end{code}

Where the client is able to do something useful with the \lstinline!Remote!
type, the server has no way to inspect \lstinline!Remote! functions, and thus
only has a no-op implementation of the \lstinline!<.>! operator. On the other
hand, it does have full access to the values and side effects produced by
the \lstinline!liftServerIO! function, which is quite simple. The side effect
is performed immediately and its result saved for later, as state holding
elements should only be initialized once.

>>>>

\begin{abstract}
We propose a new programming model for web applications which is (1)
seamless; one program and one language is used to produce code for
both client and server, (2) client-centric; the programmer takes the
viewpoint of the client that runs code on the server rather than the
other way around, (3) functional and type-safe, and (4) portable;
everything is implemented as a Haskell library that implicitly takes
care of all networking code. Our aim is to improve the painful and
error-prone experience of today's standard development methods,
in which clients and servers are coded in different languages and
communicate with each other using ad-hoc protocols.
We present the design of our library called Haste.App,
an example web application that uses it, and discuss
the implementation and the compiler technology on which it depends.
\end{abstract}

@@

For a more concrete example, let's say that we want to implement a simple
``chatbox'' component for a website, to allow visitors to discuss the site's
content in real time. Using mainstream development practices and recent
technologies such as WebSockets\ \cite{websockets}, we may come up with
something like the program in figure \ref{lst:javascript-client} for our
client program. In addition, a corresponding server program would need to be
written to handle distribution of messages among clients. We will not give such
an implementation here, as we do not believe it necessary to state the problem
at hand.

@@

To run this example, an address and a port must be provided so that the client
knows which server to contact. There are several ways of doing this:
using the GHC plugin system, through Template Haskell or by slightly altering
how program entry points are treated in a compiler or wrapper script, to name
a few. A non-intrusive method when using the GHC/Haste compiler pair would
be to pass add \lstinline!-main-is setup! to both compilers' command line
and add the \lstinline!setup! function to the source code.

@@

This will instruct the server binary to listen on the port 1111 when
started, and the client to attempt contact with that port on the local machine.
The exact mechanism chosen to provide the host and port are implementation
specific, and will in the interest of clarity not be discussed further.

@@

\begin{code}
srvHello :: Server RcptList -> Server ()
srvHello remoteRcpts = do
  recipients <- remoteRcpts
  sid <- getSessionID
  liftIO $ do
    rcptMVar <- CC.newEmptyMVar
    CC.modifyMVar recipients $ \cs ->
      return ((sid, rcptMVar):cs, ())
\end{code}

@@

runClient    :: Client () -> App Done
liftServerIO :: IO a -> App (Server a)
remote       :: Remotable a
             => a -> App (Remote a)

@@

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!remote! & \parbox[t]{5cm}{Make a server side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!remote! function to
a serializable argument.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

The workings of the \lstinline!App! monad basically hinges on the
\lstinline!Server! and \lstinline!Remote! abstract data types.
\lstinline!Server! is the monad wherein any server side code is contained, and
\lstinline!Remote! denotes functions which live on the server but can
be invoked remotely by the client. The implementation of these types and the
functions that operate on them differ between the client and the server.

\paragraph{Client side implementations}
We begin by looking at the client side implementation for those two types.

\begin{code}
data Server a = Server
data Remote a = Remote CallID [JSON]
\end{code}

The \lstinline!Server! monad is quite uninteresting to the client; since
operations performed within it can not be observed by the client in any way,
such computations are simply represented by a dummy value.
The \lstinline!Remote! type contains the identifier of a remote function and a
list of the serialized arguments to be passed when invoking it. In essence,
it is an explicit representation of a remote closure. Such closures can be
applied to values using the \lstinline!<.>! operator.

@@

The \lstinline!remote! function is used to bring server side functions into
scope on the client as \lstinline!Remote! functions. It is implemented using a
simple counter which keeps track of how many functions have been imported so
far and thus which identifier to assign to the next remote function.

@@

\begin{code}
liftServerIO :: IO a -> App (Server a)
liftServerIO _ = App Server
\end{code}

As we can see, the implementation is as simple as can be. Since
\lstinline!Server! is represented by a dummy value on the client, we
just return said value.

\paragraph{Server side implementations}
The server side representation of the \lstinline!Server! and
\lstinline!Remote! types are in a sense the opposites of their client side
counterparts.

\begin{code}
newtype Server a = Server (ReaderT SessionInfo IO a)
  deriving (Functor, Applicative, Monad, MonadIO)
data Remote a = Remote
\end{code}

Where the client is able to do something useful with the \lstinline!Remote!
type but can't touch \lstinline!Server! values, the server has no way to
inspect \lstinline!Remote! functions, and thus only has a no-op implementation
of the \lstinline!<.>! operator. On the other hand, it does have full access to
the values and side effects of the \lstinline!Server! monad, which is an IO
monad with some additional session data for the convenience of server
side code.

\lstinline!Server! values are produced by the \lstinline!liftServerIO! and
\lstinline!remote! functions. \lstinline!liftServerIO! is quite simple:
the function executes its argument immediately and the result is returned,
tucked away within the \lstinline!Server! monad.
